---
title: "Illustrative example for `hawkes` package."
author: "Peter F Halpin"
date: "5/10/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a minimal example of using the code in the `hawkes` package. The package is in very preliminary stages of development. The main functions are illustrated here. See the documentation for a full list functions and issues.

First install and load the package. 

```{r, echo = T, message = F}
devtools::install_github("peterhalpin/hawkes")
library("hawkes")
```
It comes with an example dataset, `email`, which is a list of two vectors containing the sendings times of email exchanges between an employer and employee. Time is recorded in hours and the employees start data is `t = 0`. See Halpin & DeBoeck (2013) for more details. The function `pp` takes a list of event times and turns it into an S3 object with some basic functionality

```{r}
pp_obj <- pp(email)
print(pp_obj)
plot(pp_obj)
```

The `gof` function can be used to visually inspect the data for temporal clustering (i.e., overdispersion relative to the homogeneous Poisson process). See `help(gof)` for more details. 

```{r, message = F}
gof(pp_obj)
```

The clustering might be explained using Hawkes process. To esimtate the model use the `EM` function and see `help(EM)` for a description. 

```{r, message = F}
hawkes_em <- EM(pp_obj, nstarts = 1, maxit = 20)
hawkes_em$Solutions
```

Use the gof function with the parms from the hawkes model to assess whether the model accounts for the clustering in the data. 

```{r, message = F}
parms <- get_parms(hawkes_em)
gof(pp_obj, parms)
```

Check back soon for data simulation, computation standard errors, and other functionality currently under development. 

